---
title: "Early TiC 2.0 schema adoption"
subtitle: "By Dan Snow"
format:
  html:
    fig-format: retina
    fig-dpi: 320
    toc: false
    embed-resources: true
    grid:
      body-width: 900px
execute:
  echo: false
---

```{r setup, message=FALSE}
library(arrow)
library(conflicted)
library(dplyr)
library(forcats)
library(ggplot2)
library(ggtext)
library(purrr)
library(scales)
library(stringr)
library(tidyr)

conflict_prefer_all("dplyr", quiet = TRUE)
```

```{r load_data}
DATA_DIR <- "data/output"

# Load each parquet file in data/output/, stripping the date prefix to
# create the variable name (e.g. 2026_02_18_payer_files_feb.parquet
# becomes payer_files_feb).
for (path in list.files(DATA_DIR, pattern = "\\.parquet$", full.names = TRUE)) {
  name <- sub("^\\d{4}_\\d{2}_\\d{2}_", "", tools::file_path_sans_ext(basename(path)))
  assign(
    name,
    read_parquet(path) |>
      mutate(across(any_of("payer_id"), \(x) as.character(x))) |>
      mutate(across(
        any_of("version"),
        \(x) str_extract(version, "\\d+"),
        .names = "{.col}_parsed"
      ))
  )
}

# Build a payer name lookup table w/ covered lives
payer_names <- payer_files_dec |>
  distinct(payer_id, payer_name) |>
  group_by(payer_id) |>
  slice_head(n = 1) |>
  ungroup() |>
  left_join(payer_lives, by = c("payer_id" = "tq_payer_payer_id")) |>
  select(payer_id, payer_name, payer_lives = lives)

# Grab the unique network names + count per data source/file
network_names_per_file_feb <- payer_files_feb |>
  select(payer_id, data_source_name) |>
  left_join(network_names_feb, by = c("payer_id", "data_source_name")) |>
  group_by(payer_id, data_source_name) |>
  summarize(
    nn_unique_names_lst_feb = list(unique(str_trim(flatten_chr(network_names)))),
    nn_unique_names_cnt_feb = map_int(nn_unique_names_lst_feb, \(x) length(x))
  )

# Grab the most common network name by payer + count of that name across files
network_names_ranked_per_payer_feb <- network_names_per_file_feb |>
  unnest(cols = nn_unique_names_lst_feb) |>
  ungroup() |>
  count(payer_id, nn_unique_names_lst_feb) |>
  arrange(payer_id, desc(n)) |>
  filter(row_number() <= 1, .by = payer_id) |>
  select(
    payer_id,
    nn_most_common_name_feb = nn_unique_names_lst_feb,
    nn_most_common_name_feb_file_count = n
  )

# Join various payer datasets to make our main working table
payer_df <- payer_files_dec |>
  distinct(payer_id, data_source_name, version_parsed) |>
  summarize(
    n_dsn_dec = n(),
    n_dsn_v1_dec = sum(version_parsed == 1),
    n_dsn_v2_dec = sum(version_parsed == 2),
    .by = "payer_id"
  ) |>
  left_join(
    payer_files_feb |>
      distinct(payer_id, data_source_name, version_parsed) |>
      summarize(
        n_dsn_feb = n(),
        n_dsn_v1_feb = sum(version_parsed == 1),
        n_dsn_v2_feb = sum(version_parsed == 2),
        .by = "payer_id"
      ),
    by = "payer_id"
  ) |>
  # Attach expected number of files for each month
  left_join(
    parsing_status_dec |>
      filter(!is_skipped, !is_broken, !has_error, schema == "in-network-rates") |>
      summarize(num_files_expected_dec = n(), .by = "payer_id"),
    by = "payer_id"
  ) |>
  left_join(
    parsing_status_feb |>
      filter(!is_skipped, !is_broken, !has_error, schema == "in-network-rates") |>
      summarize(num_files_expected_feb = n(), .by = "payer_id"),
    by = "payer_id"
  ) |>
  # Attach metadata (size, num external files) for parsed files
  left_join(
    parsing_status_dec |>
      filter(state == "succeed", schema == "in-network-rates") |>
      group_by(payer_id) |>
      summarize(
        num_files_parsed_dec = n(),
        num_dsn_pg_ref_ext_dec = sum(parsed_location_files > 0, na.rm = TRUE),
        tot_file_bytes_dec = sum(raw_file_size_bytes, na.rm = TRUE),
        avg_file_bytes_dec = mean(raw_file_size_bytes, na.rm = TRUE)
      ),
    by = "payer_id"
  ) |>
  left_join(
    parsing_status_feb |>
      filter(state == "succeed", schema == "in-network-rates") |>
      group_by(payer_id) |>
      summarize(
        num_files_parsed_feb = n(),
        num_dsn_pg_ref_ext_feb = sum(parsed_location_files > 0, na.rm = TRUE),
        tot_file_bytes_feb = sum(raw_file_size_bytes, na.rm = TRUE),
        avg_file_bytes_feb = mean(raw_file_size_bytes, na.rm = TRUE)
      ),
    by = "payer_id"
  ) |>
  # Diff expected vs parsed, and convert raw byte values to nice strings
  mutate(
    diff_expected_files = num_files_expected_feb - num_files_expected_dec,
    diff_parsed_files = num_files_parsed_feb - num_files_parsed_dec,
    diff_bytes = tot_file_bytes_feb - tot_file_bytes_dec,
    across(
      contains("bytes"),
      \(x) case_when(
        abs(x) >= 1099511627776 ~ paste(round(x / 1099511627776.0, 2), "TB"),
        abs(x) >= 1073741824 ~ paste(round(x / 1073741824.0, 2), "GB"),
        abs(x) >= 1048576 ~ paste(round(x / 1048576.0, 2), "MB"),
        abs(x) >= 1024 ~ paste(round(x / 1024.0, 2), "KB")
      ),
      .names = "{.col}_parsed"
    )
  ) |>
  left_join(payer_names, by = "payer_id") |>
  relocate(payer_id, payer_name, payer_lives) |>
  # Attach network_name stats
  left_join(
    network_names_per_file_feb |>
      group_by(payer_id) |>
      summarize(
        pct_file_w_network_name = mean(nn_unique_names_cnt_feb > 0, na.rm = TRUE),
        nn_unique_names_lst_feb = list(unique(flatten_chr(nn_unique_names_lst_feb))),
        nn_unique_names_cnt_feb = map_int(nn_unique_names_lst_feb, length)
      ),
    by = "payer_id"
  ) |>
  left_join(
    network_names_ranked_per_payer_feb,
    by = "payer_id"
  ) |>
  left_join(
    provider_mode_dec |>
      group_by(payer_id) |>
      summarize(
        num_dsn_pg_inline_dec = sum(provider_mode == "inline"),
        num_dsn_pg_ref_int_dec = sum(provider_mode == "reference")
      ),
    by = "payer_id"
  ) |>
  left_join(
    provider_mode_feb |>
      group_by(payer_id) |>
      summarize(
        num_dsn_pg_inline_feb = sum(provider_mode == "inline"),
        num_dsn_pg_ref_int_feb = sum(provider_mode == "reference")
          
      ),
    by = "payer_id"
  ) |>
  mutate(
    num_dsn_pg_ref_int_dec = num_dsn_pg_ref_int_dec - num_dsn_pg_ref_ext_dec,
    num_dsn_pg_ref_int_feb = num_dsn_pg_ref_int_feb - num_dsn_pg_ref_ext_feb
  ) |>
  relocate(starts_with("num_dsn_pg"), .after = everything()) |>
  relocate(starts_with("num_files"), .after = n_dsn_v2_feb) |>
  relocate(starts_with("tot_file"), .after = num_files_parsed_feb) |>
  relocate(starts_with("diff"), .after = avg_file_bytes_feb_parsed)
```

```{r plt_version_adoption_overall}
#| column: page
#| out-width: 100%
#| fig-width: 8
#| fig-height: 6.5

payer_df |>
  mutate(
    n_dsn_v1_dec = replace_values(n_dsn_v1_dec, NA ~ n_dsn_dec),
    n_dsn_v2_dec = replace_values(n_dsn_v2_dec, NA ~ 0)
  ) |>
  summarize(
    v1_dec = sum(n_dsn_v1_dec > 0 & n_dsn_v2_dec == 0, na.rm = TRUE),
    v2_dec = sum(n_dsn_v2_dec > 0 & n_dsn_v1_dec == 0 , na.rm = TRUE),
    v1_v2_dec = sum(n_dsn_v1_dec != 0 & n_dsn_v2_dec != 0, na.rm = TRUE),
    v1_feb = sum(n_dsn_v1_feb > 0 & n_dsn_v2_feb == 0, na.rm = TRUE),
    v2_feb = sum(n_dsn_v2_feb > 0 & n_dsn_v1_feb == 0 , na.rm = TRUE),
    v1_v2_feb = sum(n_dsn_v1_feb != 0 & n_dsn_v2_feb != 0, na.rm = TRUE),
    across(starts_with("v"), \(x) x / 192, .names = "pct_{.col}")
  ) |>
  rename_with(\(x) paste0("cnt_", x), starts_with("v")) |>
  pivot_longer(
    cols = everything(),
    names_to = c("stat", "version", "month"),
    names_pattern = "(pct|cnt)_(v.*)_(dec|feb)",
    values_to = "value"
  ) |>
  mutate(
    version = recode_values(
      version,
      "v1" ~ "All V1<br>files",
      "v1_v2" ~ "Mixed<br>(V1 & V2)",
      "v2" ~ "All V2<br>files"
    ),
    month = recode_values(
      month,
      "dec" ~ "December 2025",
      "feb" ~ "February 2026"
    )
  ) |>
  pivot_wider(names_from = "stat", values_from = "value") |>
  bind_rows(
    data.frame(
      version = "Missing all<br>Feb. files",
      month = "February 2026",
      cnt = 52,
      pct = 0.2708333
    )
  ) |>
  mutate(
    version = factor(
      version,
      levels = c(
        "All V1<br>files", "Mixed<br>(V1 & V2)",
        "All V2<br>files", "Missing all<br>Feb. files"
      )
    ),
    label = paste0(
      cnt, "<br>",
      "<span style='font-size:12px;color:grey30'>(",
      label_percent()(pct), ")</span>"
    )
  ) |>
ggplot() +
  geom_col(
    aes(y = cnt, x = version, fill = version)
  ) +
  geom_richtext(
    aes(y = cnt, x = version, label = label),
    nudge_y = 10,
    size = 4.5,
    label.colour = "transparent",
    fill = "transparent",
    lineheight = 0.9
  ) +
  scale_y_continuous(labels = label_comma()) +
  scale_fill_manual(
    values = c(
      "All V1<br>files" = "#fc8d62",
      "Mixed<br>(V1 & V2)" = "#8da0cb",
      "All V2<br>files" = "#66c2a5",
      "Missing all<br>Feb. files" = "grey80"
    )
  ) +
  labs(
    title = "Overall TiC 2.0 schema adoption as of Feb. 23rd, 2026",
    subtitle = paste0(
      "Of 192 payers, just 44 have fully adopted the TiC 2.0 schema ",
      "in all their files"
    ),
    caption = paste0(
      "Data from Turquoise Health payer database of in-network rate files. ",
      "Version is self-reported by payers and may not actually reflect ",
      "file changes"
    ),
    x = "",
    y = "Count of payers"
  ) +
  facet_wrap(vars(month), strip.position = "bottom", scales = "free_x") +
  guides(fill = "none") +
  theme_minimal() +
  theme(
    plot.title = element_markdown(
      size = 20,
      face = "bold",
      margin = margin(b = 8, l = -26)
    ),
    plot.subtitle = element_markdown(
      size = 15,
      margin = margin(b = 18, l = -26),
      lineheight = 1.1
    ),
    plot.caption = element_text(
      size = 9,
      lineheight = 1.1,
      margin = margin(t = 12)
    ),
    axis.title.x = element_markdown(
      margin = margin(t = 10),
      size = 16,
      face = "bold"
    ),
    axis.title.y = element_text(
      margin = margin(r = 10),
      size = 16,
      lineheight = 1.1,
      face = "bold"
    ),
    axis.text.x.bottom = element_markdown(
      size = 12,
      margin = margin(t = -8),
      lineheight = 1.3
    ),
    axis.text.y.left = element_markdown(
      size = 12,
      lineheight = 1.3
    ),
    panel.grid.major.x = element_blank(),
    panel.spacing.x = unit(1, "cm"),
    strip.placement = "outside",
    strip.text.x.bottom = element_markdown(
      size = 16,
      face = "bold",
      margin = margin(t = 10)
    )
  )
```

```{r plt_version_adoption_by_payer}
#| column: page
#| out-width: 100%
#| fig-width: 8
#| fig-height: 8

payer_df |>
  mutate(
    n_dsn_dec = replace_values(n_dsn_dec, NA ~ 0),
    n_dsn_feb = replace_values(n_dsn_feb, NA ~ 0),
    n_dsn_v2_dec = replace_values(n_dsn_v2_dec, NA ~ 0),
    n_dsn_v1_dec = pmax(
      replace_values(n_dsn_v1_dec, NA ~ n_dsn_dec),
      n_dsn_dec - n_dsn_v2_dec
    ),
    n_dsn_v1_feb = replace_values(n_dsn_v1_feb, NA ~ n_dsn_feb),
    n_dsn_v2_feb = replace_values(n_dsn_v2_feb, NA ~ 0),
    num_files_expected_dec = pmax(replace_values(
      num_files_expected_dec, NA ~ n_dsn_dec
    ), n_dsn_dec),
    num_files_expected_feb = replace_values(
      num_files_expected_feb, NA ~ num_files_expected_dec
    ),
    n_dsn_missing_dec = num_files_expected_dec - n_dsn_dec,
    n_dsn_missing_feb = num_files_expected_feb - n_dsn_feb
  ) |>
  slice_max(order_by = payer_lives, n = 40) |>
  select(
    payer_id, payer_name, payer_lives,
    starts_with("n_dsn_v"), starts_with("n_dsn_m"),
    starts_with("num_files_exp")
  ) |>
  pivot_longer(
    cols = starts_with("n_dsn"),
    names_to = c("version", "month"),
    values_to = "count",
    names_pattern = "n_dsn_(missing|v1|v2)_(dec|feb)"
  ) |>
  mutate(
    payer_name = fct_reorder(payer_name, payer_lives),
    pct = count / ifelse(
      month == "dec",
      num_files_expected_dec,
      num_files_expected_feb
    ),
    version = fct_rev(factor(
      recode_values(
        version,
        "v1" ~ "V1",
        "v2" ~ "V2",
        "missing" ~ "Missing or not yet parsed"
      ),
      levels = c("V2", "V1", "Missing or not yet parsed")
    ))
  ) |>
  filter(month != "dec") |>
  mutate(
    v2_pct = pct[version == "V2"],
    v1_pct = pct[version == "V1"],
    .by = "payer_id"
  ) |>
  arrange(desc(v2_pct), desc(v1_pct)) |>
  mutate(payer_name = fct_rev(fct_reorder(payer_name, row_number()))) |>
  ggplot() +
  geom_col(
    aes(y = payer_name, x = pct, fill = version),
    width = 0.8
  ) +
  scale_fill_manual(
    values = c(
      "V1" = "#fc8d62",
      "V2" = "#66c2a5",
      "Missing or not yet parsed" = "grey80"
    )
  ) +
  scale_x_continuous(labels = label_percent(), expand = c(0, 0)) +
  guides(fill = guide_legend(title = "", reverse = TRUE)) +
  labs(
    title = "Payer-level TiC 2.0 schema adoption as of Feb. 23rd, 2026",
    subtitle = paste0(
      "Of the top 40 payers, most are still transitioning ",
      "or have yet to post February files"
    ),
    caption = paste0(
      "Data from Turquoise Health payer database of in-network rate files. ",
      "Version is self-reported by payers and may not actually reflect ",
      "file changes"
    ),
    y = "",
    x = "Percent of payer files by type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(
      size = 19,
      face = "bold",
      margin = margin(b = 8, l = -200)
    ),
    plot.subtitle = element_markdown(
      size = 15,
      margin = margin(b = 18, l = -200),
      lineheight = 1.1
    ),
    plot.caption = element_markdown(
      size = 9,
      lineheight = 1.1,
      margin = margin(t = 12, b = 4, r = -15)
    ),
    axis.title.x = element_markdown(
      margin = margin(t = 10),
      size = 16,
      face = "bold"
    ),
    axis.title.y = element_text(
      margin = margin(r = 10),
      size = 16,
      lineheight = 1.1,
      face = "bold"
    ),
    axis.text.x.bottom = element_markdown(
      size = 12,
      lineheight = 1.3
    ),
    panel.grid.major.x = element_blank(),
    panel.spacing.x = unit(1, "cm"),
    axis.ticks.x = element_line(color = "grey40"),
    legend.position = "top",
    legend.box.margin = margin(b = -6),
    legend.margin = margin(l = -10, r = 8, t = 6, b = 6),
    legend.text = element_text(
      size = 12
    ),
    legend.title = element_text(
      size = 14,
      face = "bold",
      margin = margin(r = 20)
    ),
    legend.background = element_rect(color = "grey20"),
    plot.margin = margin(l = -20, r = 20, t = 6)
  )
```

```{r plt_pg_overall}
#| column: page
#| out-width: 100%
#| fig-width: 8
#| fig-height: 6.5

payer_df |>
  summarize(
    across(
      c("n_dsn_dec", "n_dsn_feb") | starts_with("num_dsn_pg"),
      \(x) sum(x, na.rm = TRUE)
    )
  ) |>
  pivot_longer(
    cols = starts_with("num_dsn_pg"),
    names_to = c("ref_type", "month"),
    names_pattern = "num_dsn_pg_(.*)_(dec|feb)",
    values_to = "count"
  ) |>
  mutate(pct = count / ifelse(month == "dec", n_dsn_dec, n_dsn_feb)) |>
  mutate(
    ref_type = factor(
      ref_type,
      levels = c(
        "inline",
        "ref_ext",
        "ref_int"
      )
    ),
    month = recode_values(
      month,
      "dec" ~ "December 2025",
      "feb" ~ "February 2026"
    ),
    ref_type_fct = fct_relabel(
      ref_type,
      \(x) recode_values(
        x,
        "inline" ~ paste0(
          "**Inline<br>references**<br>",
          "<span style='font-size:13px'>(allowed in 1.0)</span>"
        ),
        "ref_ext" ~ paste0(
          "**External<br>references**<br>",
          "<span style='font-size:13px'>(allowed in 1.0)</span>"
        ),
        "ref_int" ~ paste0(
          "**Internal<br>references**<br>",
          "<span style='font-size:13px'>(required in 2.0)</span>"
        )
      )
    ),
    label = label_percent()(pct)
  ) |>
ggplot() +
  geom_col(
    aes(y = pct, x = ref_type_fct, fill = ref_type)
  ) +
  geom_richtext(
    aes(y = pct, x = ref_type_fct, label = label),
    nudge_y = 0.025,
    size = 4.5,
    label.colour = "transparent",
    fill = "transparent",
    lineheight = 0.9
  ) +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_manual(
    values = c(
      "inline" = "#fc8d62",
      "ref_ext" = "#8da0cb",
      "ref_int" = "#66c2a5"
    )
  ) +
  labs(
    title = "Provider reference schema adoption as of Feb. 23rd, 2026*",
    subtitle = paste0(
      "Most payers were already using the internal ",
      "provider groups required in TiC 2.0"
    ),
    caption = paste0(
      "Data from Turquoise Health payer database of in-network rate files<br>",
      "* As a percentage of *posted* files. Not all files are posted / parsed ",
      "for February. Percentages may change as more files are added"
    ),
    x = "",
    y = "Percent of posted payer files by type"
  ) +
  facet_wrap(vars(month), strip.position = "bottom", scales = "free_x") +
  guides(fill = "none") +
  theme_minimal() +
  theme(
    plot.title = element_markdown(
      size = 19,
      face = "bold",
      margin = margin(b = 8, l = -30)
    ),
    plot.subtitle = element_markdown(
      size = 15,
      margin = margin(b = 10, l = -30),
      lineheight = 1.1
    ),
    plot.caption = element_markdown(
      size = 9,
      lineheight = 1.1,
      margin = margin(t = 6)
    ),
    axis.title.x = element_markdown(
      margin = margin(t = 10),
      size = 16,
      face = "bold"
    ),
    axis.title.y = element_text(
      margin = margin(r = 10),
      size = 16,
      lineheight = 1.1,
      face = "bold"
    ),
    axis.text.x.bottom = element_markdown(
      size = 12,
      margin = margin(t = -8),
      lineheight = 1.1
    ),
    axis.text.y.left = element_markdown(
      size = 12,
      lineheight = 1.3
    ),
    panel.grid.major.x = element_blank(),
    panel.spacing.x = unit(1, "cm"),
    strip.placement = "outside",
    strip.text.x.bottom = element_markdown(
      size = 16,
      face = "bold",
      margin = margin(t = 10)
    )
  )
```

```{r plt_pg_by_payer}
#| column: page
#| out-width: 100%
#| fig-width: 8
#| fig-height: 8

payer_df |>
  mutate(
    n_dsn_pg_all_dec = rowSums(pick(matches("num_dsn_pg_.*_dec")), na.rm = TRUE),
    n_dsn_pg_all_feb = rowSums(pick(matches("num_dsn_pg_.*_feb")), na.rm = TRUE),
    across(
      starts_with("num_dsn_pg") & ends_with("dec"),
      \(x) x / n_dsn_pg_all_dec
    ),
    across(
      starts_with("num_dsn_pg") & ends_with("feb"),
      \(x) x / n_dsn_pg_all_feb
    ),
    ref_int_diff = num_dsn_pg_ref_int_feb - num_dsn_pg_ref_int_dec
  ) |>
  slice_max(order_by = ref_int_diff, n = 36) |>
  select(
    payer_id, payer_name, payer_lives,
    starts_with("num_dsn_pg"), ref_int_diff
  ) |>
  pivot_longer(
    cols = starts_with("num_dsn_pg"),
    names_to = c("ref_type", "month"),
    names_pattern = "num_dsn_pg_(.*)_(dec|feb)",
    values_to = "pct"
  ) |>
  mutate(
    ref_type = factor(
      ref_type,
      levels = c(
        "inline",
        "ref_ext",
        "ref_int"
      )
    ),
    ref_type = fct_relabel(
      ref_type,
      \(x) recode_values(
        x,
        "inline" ~ "Inline references",
        "ref_ext" ~ "External references",
        "ref_int" ~ "Internal references"
      )
    ),
    month = recode_values(
      month,
      "dec" ~ "December 2025",
      "feb" ~ "February 2026"
    )
  ) |>
  mutate(
    inline_rank = pct[ref_type == "Inline references" & month == "December 2025"],
    external_rank = pct[ref_type == "External references" & month == "December 2025"],
    .by = "payer_id"
  ) |>
  arrange(inline_rank, desc(external_rank)) |>
  mutate(payer_name = fct_rev(fct_reorder(payer_name, row_number()))) |>
  ggplot() +
  geom_col(
    aes(y = payer_name, x = pct, fill = ref_type),
    width = 0.8
  ) +
  facet_wrap(vars(month)) +
  scale_fill_manual(
    values = c(
      "Inline references" = "#fc8d62",
      "External references" = "#8da0cb",
      "Internal references" = "#66c2a5"
    )
  ) +
  scale_x_continuous(labels = label_percent(), expand = c(0, 0)) +
  guides(fill = guide_legend(title = "", reverse = FALSE)) +
  labs(
    title = "Most major payers already use internal provider references",
    subtitle = paste0(
      "Between December and February, most remaining ",
      "payers made the swap to 2.0*"
    ),
    caption = paste0(
      "Data from Turquoise Health payer database of in-network rate files<br>",
      "* Shown as a percentage of *posted* files. Not all files are ",
      "posted / parsed for February. Percentages may change as more ",
      "files are added"
    ),
    y = "",
    x = "Percent of posted payer files by type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(
      size = 19,
      face = "bold",
      margin = margin(b = 8, l = -210)
    ),
    plot.subtitle = element_markdown(
      size = 15,
      margin = margin(b = 18, l = -210),
      lineheight = 1.1
    ),
    plot.caption = element_markdown(
      size = 9,
      lineheight = 1.1,
      margin = margin(t = 16, b = 4, r = -15)
    ),
    axis.title.x = element_markdown(
      margin = margin(t = 10),
      size = 16,
      face = "bold"
    ),
    axis.title.y = element_text(
      margin = margin(r = 10),
      size = 16,
      lineheight = 1.1,
      face = "bold"
    ),
    axis.text.x.bottom = element_markdown(
      size = 12,
      lineheight = 1.3
    ),
    panel.grid.major.x = element_blank(),
    panel.spacing.x = unit(1, "cm"),
    axis.ticks.x = element_line(color = "grey40"),
    strip.text.x.top = element_text(
      size = 14,
      face = "bold"
    ),
    legend.position = "top",
    legend.box.margin = margin(b = -6, l = -200),
    legend.margin = margin(l = -10, r = 8, t = 6, b = 6),
    legend.text = element_text(
      size = 12
    ),
    legend.title = element_text(
      size = 14,
      face = "bold",
      margin = margin(r = 20)
    ),
    legend.background = element_rect(color = "grey20"),
    plot.margin = margin(l = -20, r = 20, t = 6)
  )
```

```{r plt_nn_by_payer}
#| column: page
#| out-width: 100%
#| fig-width: 8
#| fig-height: 6.5

payer_df |>
  filter(pct_file_w_network_name > 0) |>
  ggplot() +
  geom_point(
    aes(
      x = pct_file_w_network_name,
      y = uniq_network_name_cnt,
      size = payer_lives
    ),
    alpha = 0.7
  ) +
  scale_x_continuous(
    name = "Percent of posted files with network name",
    labels = label_percent()
  ) +
  scale_y_log10(
    name = "Number of unique network names (log)",
    labels = label_comma()
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(
      size = 19,
      face = "bold",
      margin = margin(b = 8, l = -210)
    ),
    plot.subtitle = element_markdown(
      size = 15,
      margin = margin(b = 18, l = -210),
      lineheight = 1.1
    ),
    plot.caption = element_markdown(
      size = 9,
      lineheight = 1.1,
      margin = margin(t = 16, b = 4, r = -15)
    ),
    axis.title.x = element_markdown(
      margin = margin(t = 10),
      size = 16,
      face = "bold"
    ),
    axis.title.y = element_text(
      margin = margin(r = 10),
      size = 16,
      lineheight = 1.1,
      face = "bold"
    ),
    axis.text.x = element_markdown(
      size = 12
    ),
    axis.text.y = element_markdown(
      size = 12
    ),
    legend.position = "inside",
    legend.box.margin = margin(b = -6, l = -200),
    legend.margin = margin(l = -10, r = 8, t = 6, b = 6),
    legend.text = element_text(
      size = 12
    ),
    legend.title = element_text(
      size = 14,
      face = "bold",
      margin = margin(r = 20)
    ),
    legend.background = element_rect(color = "grey20")
  )

```


---

```{r network_names_plot_pct_payer}
payer_df |>
  summarize(
    n_payers = n(),
    n_payers_w_nn = sum(uniq_network_name_cnt > 0, na.rm = TRUE),
    pct_payers_w_nn = n_payers_w_nn / n_payers,
    tot_lives = sum(payer_lives, na.rm = TRUE),
    tot_lives_w_nn = sum(payer_lives * (uniq_network_name_cnt > 0), na.rm = TRUE),
    pct_tot_lives_w_nn = tot_lives_w_nn / tot_lives
  ) |>
ggplot() +
  geom_col(
    aes(x = "Percent of payers", y = 1),
    fill = "grey50"
  ) +
  geom_col(
    aes(x = "Percent of payers", y = pct_payers_w_nn),
    fill = "lightblue"
  ) +
  geom_col(
    aes(x = "Percent of lives", y = 1),
    fill = "grey50"
  ) +
  geom_col(
    aes(x = "Percent of lives", y = pct_tot_lives_w_nn),
    fill = "lightblue"
  )
```

```{r}
temp <- parsing_status_dec |>
  filter(!is_broken, !has_error, !is_skipped, schema == "in-network-rates") |>
  anti_join(
    payer_files_dec |> select(payer_id, data_source_name),
    by = c("payer_id", "data_source_name")
  )

temp <- provider_mode_dec |>
  filter(provider_mode == "inline") |>
  select(payer_id, dsn_dec = data_source_name) |>
  left_join(
    parsing_status_dec |>
      select(payer_id, data_source_name, bytes_dec = raw_file_size_bytes),
    by = c("payer_id", "dsn_dec" = "data_source_name")
  ) |>
  mutate(
  ) |>
  left_join(
    provider_mode_feb |>
      filter(provider_mode == "reference") |>
      select(payer_id, dsn_feb = data_source_name) |>
      left_join(
        parsing_status_feb |>
          select(payer_id, data_source_name, bytes_feb = raw_file_size_bytes),
        by = c("payer_id", "dsn_feb" = "data_source_name")
      ) |>
      mutate(
      ),
    by = c("payer_id", "dsn_dec" = "dsn_feb")
  )


temp2 <- temp |>
  group_by(data_source_name_dec) |>
  filter(dsn_dist == min(dsn_dist))

temp <- network_names_per_file |>
  group_by(payer_name, payer_id, lives) |>
  summarize(
    file_count = n(),
    pct_w_network_name = mean(n_uniq > 0),
  ) |>
  ungroup()
  summarize(
    cnt_payers_w_network_name = sum(pct_w_network_name > 0),
    pct_payers_w_network_name = mean(pct_w_network_name > 0)
  )
```

CMS updated the Transparency in Coverage (TiC) machine-readable file schema
to version 2.0, effective February 2, 2026. The update includes several
categories of changes. Three affect the provider data model directly:

1. **`network_name` required** in provider references
2. **Provider groups consolidated** into mandatory in-file provider references (eliminating both inline groups and external reference files)
3. **`business_name` required** alongside EIN records

The v2 schema also makes `setting` required on negotiated prices and
`issuer_name` required in plan metadata; `severity_of_illness` is added as an
optional field. This analysis focuses on the provider-level structural changes.

We looked at early February 2026 data to measure how quickly payers are
adopting these changes. This analysis uses raw machine-readable file data, not
TQ's compressed schema (which incorporates the new v2 fields starting with the
March 13 release). A few caveats:

- **February ingestion is still in progress.** Our February data covers
  `r n_distinct(payer_files_feb$payer_id)` payers compared to
  `r n_distinct(payer_files_dec$payer_id)` in December 2025. Some payers haven't
  been ingested yet, so adoption rates will shift.
- **The v2 effective date is February 2, not February 1.** Payers that posted
  files dated 2/1/2026 using v1 are technically compliant. Some payers posted
  v2 files as early as January.
- **Version is self-reported.** A payer claiming v2 in the `version` field
  doesn't guarantee their files actually conform to the new schema.
- **v2.x includes minor revisions.** The v2.0.0 schema was finalized October
  1, 2025; v2.2.0 followed on February 2, 2026 with relaxed constraints for
  allowed-amounts files. This analysis groups all v2.x versions together.

```{r prep-version}
payer_ver <- payer_files_feb |>
  mutate(
    version_major = case_when(
      str_starts(version, "2") | str_starts(str_to_lower(version), "v2") ~ "v2",
      str_starts(version, "1") | str_starts(str_to_lower(version), "v1") ~ "v1",
      TRUE ~ "missing"
    )
  )

payer_ver_summary <- payer_ver |>
  group_by(payer_id) |>
  summarise(
    has_v2 = any(version_major == "v2"),
    all_v2 = all(version_major == "v2"),
    n_files = n(),
    .groups = "drop"
  ) |>
  mutate(
    ver_class = case_when(
      all_v2 ~ "All v2",
      has_v2 ~ "Mixed (v1 + v2)",
      TRUE ~ "All v1"
    )
  )

n_payers <- nrow(payer_ver_summary)
n_v2_any <- sum(payer_ver_summary$has_v2)
n_v2_all <- sum(payer_ver_summary$all_v2)
n_v1_only <- sum(!payer_ver_summary$has_v2)
```

## Self-reported version

Of the **`r n_payers`** payers submitting files in February 2026,
**`r n_v2_any`** (`r round(n_v2_any / n_payers * 100)`%) report at least one
v2 file. Only **`r n_v2_all`** have moved all files to v2; another
**`r n_v2_any - n_v2_all`** are in a mixed state with some files still on v1.
The remaining **`r n_v1_only`** payers report only v1.

```{r plot-version, fig.width=7, fig.height=4}
ver_payer_counts <- payer_ver_summary |>
  count(ver_class, name = "n_payers") |>
  mutate(
    ver_class = factor(
      ver_class,
      levels = c("All v1", "Mixed (v1 + v2)", "All v2")
    )
  )

ggplot(ver_payer_counts, aes(x = ver_class, y = n_payers)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(
    aes(label = n_payers),
    vjust = -0.5,
    size = 4.5,
    fontface = "bold"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Payer-level version adoption in February 2026",
    subtitle = paste0(
      "Of ",
      n_payers,
      " payers, ",
      n_v2_any,
      " report at least one v2 file"
    ),
    x = NULL,
    y = "Number of payers",
    caption = paste0(
      "Source: hive.public_2026_02.inr\n",
      "A payer is classified by the version field across all its files."
    )
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(
      size = 18,
      face = "bold",
      margin = margin(b = 6, l = -36)
    ),
    plot.subtitle = element_markdown(
      size = 15,
      lineheight = 1.1,
      margin = margin(b = 12, l = -36)
    ),
    plot.caption = element_text(
      size = 11,
      lineheight = 1.1,
      margin = margin(t = 16)
    ),
    axis.text.x = element_markdown(size = 14, face = "bold"),
    axis.text.y = element_markdown(size = 14),
    axis.title.y = element_text(
      size = 16,
      face = "bold",
      margin = margin(r = 10)
    ),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Self-reported version doesn't guarantee actual schema compliance. The
sections below examine whether payers are actually implementing the three
v2 structural requirements.

```{r prep-provider-mode}
classify_modes <- function(df) {
  df |>
    distinct(payer_id, provider_mode) |>
    group_by(payer_id) |>
    summarise(modes = list(provider_mode), .groups = "drop") |>
    mutate(
      has_inline = sapply(modes, \(x) "inline" %in% x),
      has_ref = sapply(modes, \(x) "reference" %in% x),
      mode_class = case_when(
        has_inline & has_ref ~ "Both",
        has_inline ~ "Inline only",
        TRUE ~ "References only"
      )
    )
}

mode_feb <- classify_modes(provider_mode_feb)
mode_dec <- classify_modes(provider_mode_dec)

# Transition matrix: Dec mode -> Feb mode for payers in both periods
mode_transitions <- mode_dec |>
  select(payer_id, dec_mode = mode_class) |>
  inner_join(
    mode_feb |> select(payer_id, feb_mode = mode_class),
    by = "payer_id"
  )

transition_counts <- mode_transitions |>
  count(dec_mode, feb_mode, name = "n_payers") |>
  mutate(
    dec_mode = factor(
      dec_mode,
      levels = c("Inline only", "Both", "References only")
    ),
    feb_mode = factor(
      feb_mode,
      levels = c("Inline only", "Both", "References only")
    )
  )

# Fill in any missing combos with 0
transition_counts <- transition_counts |>
  complete(dec_mode, feb_mode, fill = list(n_payers = 0))

n_inline_dec <- sum(mode_dec$mode_class == "Inline only")
n_inline_feb <- sum(mode_feb$mode_class == "Inline only")

# Payers that were inline-only in Dec and are now in both periods
inline_dec_payers <- mode_transitions |>
  filter(dec_mode == "Inline only")
n_transitioned <- sum(inline_dec_payers$feb_mode != "Inline only")
n_still_inline <- sum(inline_dec_payers$feb_mode == "Inline only")

# Name the payers that transitioned
transitioned_payers <- inline_dec_payers |>
  filter(feb_mode != "Inline only") |>
  left_join(payer_names, by = "payer_id")
```

## Provider references vs. inline groups

The v2 schema eliminates both inline provider groups and external
provider-reference files in favor of in-file provider references, a structural
change that reduces file duplication and file size.
Rather than comparing raw counts across periods (which differ in the number
of payers ingested), we track individual payer transitions.

Of the **`r nrow(mode_transitions)`** payers present in both December 2025
and February 2026, the heatmap below shows how each payer's provider mode
changed:

```{r plot-provider-transition, fig.width=7, fig.height=5}
ggplot(
  transition_counts,
  aes(x = feb_mode, y = dec_mode, fill = n_payers)
) +
  geom_tile(color = "white", linewidth = 1.5) +
  geom_text(
    aes(label = n_payers),
    size = 6,
    fontface = "bold"
  ) +
  scale_fill_gradient(low = "grey95", high = "steelblue", guide = "none") +
  labs(
    title = "Provider mode transitions: December 2025 to February 2026",
    subtitle = paste0(
      nrow(mode_transitions),
      " payers present in both periods"
    ),
    x = "February 2026",
    y = "December 2025",
    caption = paste0(
      "Source: inr_in_network_neg_rates_provider_groups ",
      "and inr_in_network_neg_rates_provider_references"
    )
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(
      size = 18,
      face = "bold",
      margin = margin(b = 6, l = -36)
    ),
    plot.subtitle = element_markdown(
      size = 15,
      lineheight = 1.1,
      margin = margin(b = 12, l = -36)
    ),
    plot.caption = element_text(
      size = 11,
      lineheight = 1.1,
      margin = margin(t = 16)
    ),
    axis.text.x = element_markdown(size = 14, face = "bold"),
    axis.text.y = element_markdown(size = 14, face = "bold"),
    axis.title = element_text(
      size = 16,
      face = "bold"
    ),
    panel.grid = element_blank()
  )
```

Most payers already used provider references before v2. Of the
**`r n_inline_dec`** payers that used inline groups exclusively in December,
**`r n_transitioned`** have transitioned to references by February, while
**`r n_still_inline`** remain inline-only.

```{r transitioned-names, results='asis'}
if (nrow(transitioned_payers) > 0 && nrow(transitioned_payers) <= 20) {
  cat(
    "Payers that transitioned away from inline-only: ",
    paste0(
      "**",
      transitioned_payers$payer_name,
      "**",
      collapse = ", "
    ),
    ".\n"
  )
}
```

```{r load-file-sizes}
file_sizes_dec <- read_parquet(paste0(
  "data/output/", DATA_DATE, "_file_sizes_dec.parquet"
)) |>
  group_by(payer_id) |>
  summarise(
    size_dec = sum(total_size),
    n_files_dec = sum(n_files),
    .groups = "drop"
  )

file_sizes_feb <- read_parquet(paste0(
  "data/output/", DATA_DATE, "_file_sizes_feb.parquet"
)) |>
  group_by(payer_id) |>
  summarise(
    size_feb = sum(total_size),
    n_files_feb = sum(n_files),
    .groups = "drop"
  )

payer_sizes <- file_sizes_dec |>
  inner_join(file_sizes_feb, by = "payer_id") |>
  mutate(
    pct_change = (size_feb - size_dec) / size_dec,
    size_dec_gb = size_dec / 1e9,
    size_feb_gb = size_feb / 1e9
  ) |>
  left_join(payer_names, by = "payer_id")

n_payers_sized <- nrow(payer_sizes)
n_shrank <- sum(payer_sizes$pct_change < -0.01)
n_grew <- sum(payer_sizes$pct_change > 0.01)
```

### File size changes

Moving provider information from inline groups to a separate
`provider_references` block should reduce file sizes by eliminating
duplication. We compare total raw file sizes per payer between December 2025
and February 2026 using S3 object metadata (no file downloads needed).

Of the **`r n_payers_sized`** payers present in both periods,
**`r n_shrank`** saw their total file size decrease by more than 1%, while
**`r n_grew`** saw an increase of more than 1%. The rest remained roughly
unchanged.

```{r plot-file-sizes, fig.width=8, fig.height=6}
payer_sizes_top <- payer_sizes |>
  mutate(abs_change = abs(size_feb - size_dec)) |>
  slice_max(abs_change, n = 20) |>
  mutate(
    name_label = str_trunc(
      coalesce(payer_name, payer_id),
      35
    ),
    name_label = fct_reorder(name_label, pct_change)
  )

ggplot(payer_sizes_top, aes(x = pct_change, y = name_label)) +
  geom_col(
    aes(fill = pct_change < 0),
    width = 0.7,
    show.legend = FALSE
  ) +
  geom_text(
    aes(
      label = paste0(
        ifelse(pct_change > 0, "+", ""),
        round(pct_change * 100),
        "%"
      ),
      hjust = ifelse(pct_change < 0, 1.15, -0.15)
    ),
    size = 3.5,
    fontface = "bold"
  ) +
  scale_x_continuous(
    labels = label_percent(),
    expand = expansion(mult = c(0.15, 0.15))
  ) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "coral")) +
  geom_vline(xintercept = 0, linewidth = 0.4, color = "grey30") +
  labs(
    title = "Raw file size change: December 2025 to February 2026",
    subtitle = "Top 20 payers by absolute size change (payers in both periods)",
    x = "% change in total file size",
    y = NULL,
    caption = paste0(
      "Source: S3 object metadata for in-network-rates files\n",
      "Covers ",
      n_payers_sized,
      " payers present in both Dec 2025 and Feb 2026."
    )
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(
      size = 18,
      face = "bold",
      margin = margin(b = 6, l = -36)
    ),
    plot.subtitle = element_markdown(
      size = 15,
      lineheight = 1.1,
      margin = margin(b = 12, l = -36)
    ),
    plot.caption = element_text(
      size = 11,
      lineheight = 1.1,
      margin = margin(t = 16)
    ),
    axis.text.y = element_markdown(size = 12),
    axis.text.x = element_markdown(size = 14),
    axis.title.x = element_text(
      size = 16,
      face = "bold",
      margin = margin(t = 10)
    ),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r prep-network-names}
# The query returns network_names as an array column. Compute count and
# unnest for downstream use.
network_names <- network_names |>
  mutate(n_distinct_names = lengths(network_names))

nn_with_names <- network_names |>
  filter(n_distinct_names > 0)
nn_payers_with <- n_distinct(nn_with_names$payer_id)
total_feb_payers <- n_distinct(payer_files_feb$payer_id)

# Unnest for per-name analysis
nn_long <- nn_with_names |>
  tidyr::unnest(network_names)

# Quality: how many use generic "UNSPECIFIED_NETWORK"?
nn_unspec_names <- nn_long |>
  filter(str_detect(network_names, "(?i)unspecified"))
nn_real_names <- nn_long |>
  filter(!str_detect(network_names, "(?i)unspecified"))
nn_real_payers <- n_distinct(nn_real_names$payer_id)

# Multiple networks per file
nn_multi <- network_names |> filter(n_distinct_names > 1)
nn_pct_multi <- round(
  n_distinct(nn_multi$payer_id) / n_distinct(network_names$payer_id) * 100
)
nn_median <- median(network_names$n_distinct_names)
```

## Network names

The v2 schema requires a `network_name` array in provider references,
enabling consumers to identify which provider network applies to each rate.

### Who is posting?

Of the **`r n_distinct(network_names$payer_id)`** payers with provider
references in February, **`r nn_payers_with`**
(`r round(nn_payers_with / n_distinct(network_names$payer_id) * 100)`%)
have at least one non-empty `network_name`.

### Name quality

Quality varies. Some payers provide market-legible names
(e.g., "SELECTHEALTH MED", "BLUE ADVANTAGE HMO"), while others use generic
placeholders like `UNSPECIFIED_NETWORK`. Excluding placeholders,
**`r nn_real_payers`** payers provide meaningful network names.

```{r plot-network-names, fig.width=8, fig.height=5}
nn_top <- nn_real_names |>
  count(network_names, name = "n_files", sort = TRUE) |>
  slice_head(n = 15) |>
  mutate(
    network_names = fct_reorder(network_names, n_files)
  )

ggplot(nn_top, aes(x = n_files, y = network_names)) +
  geom_col(fill = "steelblue", width = 0.7) +
  geom_text(
    aes(label = comma(n_files)),
    hjust = -0.15,
    size = 3.5,
    fontface = "bold"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = "Most common network names in February 2026",
    subtitle = paste0(
      "Top 15 names by number of payer/file combinations (excluding placeholders)"
    ),
    x = "Number of payer/file combinations",
    y = NULL,
    caption = paste0(
      "Source: hive.public_2026_02.inr_provider_references\n",
      "Counts each distinct network name per payer/file combination."
    )
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(
      size = 18,
      face = "bold",
      margin = margin(b = 6, l = -36)
    ),
    plot.subtitle = element_markdown(
      size = 15,
      lineheight = 1.1,
      margin = margin(b = 12, l = -36)
    ),
    plot.caption = element_text(
      size = 11,
      lineheight = 1.1,
      margin = margin(t = 16)
    ),
    axis.text.y = element_markdown(size = 12),
    axis.text.x = element_markdown(size = 14),
    axis.title.x = element_text(
      size = 16,
      face = "bold",
      margin = margin(t = 10)
    ),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

### Multiple networks per file?

The median number of distinct network names per payer/file is
**`r nn_median`**. **`r nn_pct_multi`%** of payers have at least one file
with more than one distinct network name. This matters practically: TQ plans
to use `network_name` to supplement file labels, starting with files that have
a single name. Files with multiple names or no name won't benefit from this
labeling.

```{r plot-network-hist, fig.width=7, fig.height=4}
nn_hist_data <- network_names |>
  mutate(
    name_bucket = case_when(
      n_distinct_names == 0 ~ "0 (none)",
      n_distinct_names == 1 ~ "1",
      n_distinct_names <= 5 ~ "2-5",
      n_distinct_names <= 20 ~ "6-20",
      TRUE ~ "20+"
    ),
    name_bucket = factor(
      name_bucket,
      levels = c("0 (none)", "1", "2-5", "6-20", "20+")
    )
  ) |>
  count(name_bucket, name = "n_files")

ggplot(nn_hist_data, aes(x = name_bucket, y = n_files)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(
    aes(label = comma(n_files)),
    vjust = -0.5,
    size = 4,
    fontface = "bold"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Distinct network names per payer/file",
    subtitle = "Distribution across all February 2026 payer/file combinations",
    x = "Number of distinct network names",
    y = "Number of payer/file combinations",
    caption = "Source: hive.public_2026_02.inr_provider_references"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(
      size = 18,
      face = "bold",
      margin = margin(b = 8, l = -36)
    ),
    plot.subtitle = element_markdown(
      size = 15,
      lineheight = 1.1,
      margin = margin(b = 12, l = -36)
    ),
    plot.caption = element_text(
      size = 11,
      lineheight = 1.1,
      margin = margin(t = 16)
    ),
    axis.text.x = element_markdown(size = 14, face = "bold"),
    axis.text.y = element_markdown(size = 14),
    axis.title.x = element_text(
      size = 16,
      face = "bold",
      margin = margin(t = 10)
    ),
    axis.title.y = element_text(
      size = 16,
      face = "bold",
      margin = margin(r = 10)
    ),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```

### Fewer files overall?

One expected benefit of provider references and network names is
consolidation: payers can put multiple networks in a single file rather than
publishing separate files per network. We compare per-payer file counts
between December 2025 and February 2026 for payers present in both periods.

```{r plot-file-counts, fig.width=8, fig.height=6}
file_counts_dec <- payer_files_dec |>
  count(payer_id, name = "n_files_dec")
file_counts_feb <- payer_files_feb |>
  count(payer_id, name = "n_files_feb")

file_compare <- file_counts_dec |>
  inner_join(file_counts_feb, by = "payer_id") |>
  left_join(payer_names, by = "payer_id") |>
  mutate(
    change = n_files_feb - n_files_dec,
    pct_change = (n_files_feb - n_files_dec) / n_files_dec * 100
  )

# Show top 20 payers by Dec file count
file_top <- file_compare |>
  slice_max(n_files_dec, n = 20) |>
  pivot_longer(
    cols = c(n_files_dec, n_files_feb),
    names_to = "period",
    values_to = "n_files"
  ) |>
  mutate(
    period = if_else(period == "n_files_dec", "Dec 2025", "Feb 2026"),
    period = factor(period, levels = c("Dec 2025", "Feb 2026")),
    name_label = str_trunc(payer_name, 35),
    name_label = fct_reorder(name_label, n_files, .fun = max)
  )

ggplot(file_top, aes(x = n_files, y = name_label, fill = period)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  scale_fill_manual(
    values = c("Dec 2025" = "grey65", "Feb 2026" = "steelblue")
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.05)),
    labels = comma
  ) +
  labs(
    title = "Per-payer file counts: December 2025 vs. February 2026",
    subtitle = "Top 20 payers by December file count (payers in both periods)",
    x = "Number of files",
    y = NULL,
    fill = NULL,
    caption = paste0(
      "Source: hive.public_2026_02.inr and hive.public_2025_12.inr\n",
      "Only includes ",
      nrow(file_compare),
      " payers present in both periods."
    )
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(
      size = 18,
      face = "bold",
      margin = margin(b = 8, l = -36)
    ),
    plot.subtitle = element_markdown(
      size = 15,
      lineheight = 1.1,
      margin = margin(b = 12, l = -36)
    ),
    plot.caption = element_text(
      size = 11,
      lineheight = 1.1,
      margin = margin(t = 16)
    ),
    axis.text.y = element_markdown(size = 12),
    axis.text.x = element_markdown(size = 14),
    axis.title.x = element_text(
      size = 16,
      face = "bold",
      margin = margin(t = 10)
    ),
    axis.ticks.y = element_blank(),
    legend.position = "top",
    legend.text = element_text(size = 12),
    legend.background = element_rect(color = "black"),
    legend.key.spacing.x = unit(0.7, "cm"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r prep-ein}
# Per-payer fill rate
ein_fill_payer <- ein_fill |>
  group_by(payer_id) |>
  summarise(
    n_ein_records = sum(n_ein_records),
    n_with_business_name = sum(n_with_business_name),
    .groups = "drop"
  ) |>
  mutate(fill_pct = n_with_business_name / n_ein_records * 100)

# From ein_agg: payer-level has/hasn't
ein_payer <- ein_agg |>
  group_by(payer_id) |>
  summarise(
    has_business_name = any(has_any_business_name),
    .groups = "drop"
  )

ein_total <- nrow(ein_payer)
ein_with <- sum(ein_payer$has_business_name)
```

## EIN business names

The v2 schema requires a `business_name` for EIN-type TINs in the `tin`
struct, making it possible to identify providers by name rather than just tax
ID number. Because `business_name` is a free-text field, the values will
require standardization before they can be used systematically for provider
enrichment.

### Adoption

Of the **`r ein_total`** payers with EIN data in February 2026,
**`r ein_with`** (`r round(ein_with / ein_total * 100)`%) include at least
one non-empty `business_name`.

### Fill rate distribution

Having *some* business names is different from filling them consistently.
The histogram below shows per-payer fill rates (what fraction of a payer's
EIN records actually include a `business_name`). The distribution is
bimodal: most payers either fill nearly all their EIN records or none.

```{r plot-ein-fill, fig.width=7, fig.height=4}
ggplot(ein_fill_payer, aes(x = fill_pct)) +
  geom_histogram(
    binwidth = 5,
    fill = "steelblue",
    color = "white",
    boundary = 0
  ) +
  scale_x_continuous(
    breaks = seq(0, 100, 25),
    labels = \(x) paste0(x, "%")
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "EIN business name fill rate by payer",
    subtitle = paste0(
      "Distribution across ",
      nrow(ein_fill_payer),
      " payers with EIN data in February 2026"
    ),
    x = "% of EIN records with a business_name",
    y = "Number of payers",
    caption = "Source: hive.public_2026_02.inr_provider_references_provider_groups"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(
      size = 18,
      face = "bold",
      margin = margin(b = 8, l = -36)
    ),
    plot.subtitle = element_markdown(
      size = 15,
      lineheight = 1.1,
      margin = margin(b = 12, l = -36)
    ),
    plot.caption = element_text(
      size = 11,
      lineheight = 1.1,
      margin = margin(t = 16)
    ),
    axis.text.x = element_markdown(size = 14),
    axis.text.y = element_markdown(size = 14),
    axis.title.x = element_text(
      size = 16,
      face = "bold",
      margin = margin(t = 10)
    ),
    axis.title.y = element_text(
      size = 16,
      face = "bold",
      margin = margin(r = 10)
    ),
    panel.grid.minor = element_blank()
  )
```

### Sample names

Are the business names useful? The table below shows a sample of EIN records
with populated `business_name` values:

```{r table-ein-sample}
ein_sample |>
  select(
    Payer = payer_name,
    EIN = tin_value,
    `Business Name` = business_name
  ) |>
  mutate(Payer = str_trunc(Payer, 40)) |>
  slice_sample(n = min(15, nrow(ein_sample))) |>
  knitr::kable()
```

```{r prep-summary}
# Build a unified payer-level summary
payer_all <- payer_ver_summary |>
  select(payer_id, has_v2) |>
  left_join(
    mode_feb |> select(payer_id, mode_class),
    by = "payer_id"
  ) |>
  left_join(
    network_names |>
      filter(n_distinct_names > 0) |>
      distinct(payer_id) |>
      mutate(has_network_name = TRUE),
    by = "payer_id"
  ) |>
  left_join(
    ein_payer |> rename(has_business_name_val = has_business_name),
    by = "payer_id"
  ) |>
  mutate(
    has_network_name = coalesce(has_network_name, FALSE),
    uses_references = mode_class %in% c("References only", "Both"),
    has_business_name_val = coalesce(has_business_name_val, FALSE)
  )

summary_tbl <- tibble(
  Requirement = c(
    "Self-reports v2",
    "Uses provider references",
    "Has network_name",
    "Has EIN business_name"
  ),
  Adopted = c(
    sum(payer_all$has_v2),
    sum(payer_all$uses_references, na.rm = TRUE),
    sum(payer_all$has_network_name),
    sum(payer_all$has_business_name_val)
  ),
  Total = n_payers,
  Notes = c(
    "Self-reported; doesn't guarantee compliance",
    "Most already used references pre-v2",
    "Quality varies; some placeholders",
    "Of payers with EIN data; bimodal fill"
  )
) |>
  mutate(
    `%` = paste0(round(Adopted / Total * 100), "%")
  ) |>
  select(Requirement, Adopted, Total, `%`, Notes)
```

## Summary

```{r table-summary}
knitr::kable(summary_tbl)
```

Provider references have the highest adoption rate --- most payers were
already using them before the v2 mandate. Network names and EIN business
names, both new v2 fields, show adoption above 50%, though the
`business_name` denominator only counts payers with EIN data
(`r ein_total` of `r n_payers`).

Looking ahead: the **`r n_v1_only`** payers still on v1-only files need to
transition before CMS enforcement tightens. Network name quality needs
improvement --- placeholder values like `UNSPECIFIED_NETWORK` don't help
consumers identify provider networks. And while early file size data shows
some payers achieving significant reductions from the v2 provider reference
structure, most payers' total sizes haven't changed meaningfully yet.
