---
title: "Early v2 TiC Schema Adoption"
subtitle: "By Dan Snow"
format:
  html:
    fig-format: retina
    fig-dpi: 320
    toc: false
    embed-resources: true
    grid:
      body-width: 900px
execute:
  echo: false
---

```{r setup, message=FALSE}
library(arrow)
library(conflicted)
library(dplyr)
library(forcats)
library(ggplot2)
library(ggtext)
library(scales)
library(stringr)
library(tidyr)

conflict_prefer_all("dplyr", quiet = TRUE)
```

```{r load-data}
payer_feb <- read_parquet("data/output/payer_files_feb.parquet")
payer_dec <- read_parquet("data/output/payer_files_dec.parquet")
network_names <- read_parquet("data/output/network_names_feb.parquet")
provider_mode_feb <- read_parquet("data/output/provider_mode_feb.parquet")
provider_mode_dec <- read_parquet("data/output/provider_mode_dec.parquet")
ein_agg <- read_parquet("data/output/ein_names_agg_feb.parquet")
```

CMS updated the Transparency in Coverage (TiC) machine-readable file schema
to version 2.0, effective February 2, 2026. The key structural changes are:

1. **`network_name` required** in provider references
2. **Inline provider groups replaced** by mandatory provider references
3. **`business_name` required** alongside EIN records

We compare December 2025 (pre-v2) to February 2026 (first month under v2)
to measure early adoption across these three dimensions.

```{r prep-version}
# Normalize version strings into major buckets
payer_ver <- payer_feb |>
  mutate(
    version_major = case_when(
      str_starts(version, "2") | str_starts(str_to_lower(version), "v2") ~ "v2",
      str_starts(version, "1") | str_starts(str_to_lower(version), "v1") ~ "v1",
      TRUE ~ "missing"
    )
  )

# Payer-level version summary
payer_ver_summary <- payer_ver |>
  group_by(payer_id) |>
  summarise(
    has_v2 = any(version_major == "v2"),
    all_v2 = all(version_major == "v2"),
    n_files = n(),
    .groups = "drop"
  ) |>
  mutate(
    ver_class = case_when(
      all_v2 ~ "All v2",
      has_v2 ~ "Mixed (v1 + v2)",
      TRUE ~ "All v1"
    )
  )

n_payers <- nrow(payer_ver_summary)
n_v2_any <- sum(payer_ver_summary$has_v2)
n_v2_all <- sum(payer_ver_summary$all_v2)
n_v1_only <- sum(!payer_ver_summary$has_v2)
```

## Self-reported version

The `version` field in the TiC schema lets payers declare which spec version
they're using. Of the **`r n_payers`** payers submitting files in February
2026, **`r n_v2_any`** (`r round(n_v2_any / n_payers * 100)`%) report at
least one v2 file. But only **`r n_v2_all`** have moved all files to v2;
another **`r n_v2_any - n_v2_all`** are in a mixed state, with some files
still on v1.

```{r plot-version, fig.width=7, fig.height=4}
ver_file_counts <- payer_ver |>
  count(version_major, name = "n_files") |>
  mutate(
    version_major = factor(
      version_major,
      levels = c("v1", "v2", "missing"),
      labels = c("v1.x", "v2.x", "Missing")
    )
  )

ver_payer_counts <- payer_ver_summary |>
  count(ver_class, name = "n_payers") |>
  mutate(
    ver_class = factor(
      ver_class,
      levels = c("All v1", "Mixed (v1 + v2)", "All v2")
    )
  )

ggplot(ver_payer_counts, aes(x = ver_class, y = n_payers)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(
    aes(label = n_payers),
    vjust = -0.5,
    size = 4.5,
    fontface = "bold"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Payer-level version adoption in February 2026",
    subtitle = paste0(
      "Of ", n_payers, " payers, ",
      n_v2_any, " report at least one v2 file"
    ),
    x = NULL,
    y = "Number of payers",
    caption = paste0(
      "Source: hive.public_2026_02.inr\n",
      "A payer is classified by the version field across all its files."
    )
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(size = 16, face = "bold", margin = margin(b = 4)),
    plot.subtitle = element_markdown(size = 13, margin = margin(b = 12)),
    plot.caption = element_text(size = 10, lineheight = 1.1, margin = margin(t = 12)),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 11),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```

The file-level picture is similar: of **`r comma(nrow(payer_feb))`** total
files in February, **`r comma(sum(ver_file_counts$n_files[ver_file_counts$version_major == "v2.x"]))`**
(`r round(sum(ver_file_counts$n_files[ver_file_counts$version_major == "v2.x"]) / nrow(payer_feb) * 100)`%)
self-report as v2.

But self-reported version alone doesn't tell us whether payers are actually
implementing the new schema features. We next look at adoption of each v2
structural requirement.

```{r prep-provider-mode}
classify_modes <- function(df) {
  df |>
    distinct(payer_id, provider_mode) |>
    group_by(payer_id) |>
    summarise(modes = list(provider_mode), .groups = "drop") |>
    mutate(
      has_inline = sapply(modes, \(x) "inline" %in% x),
      has_ref = sapply(modes, \(x) "reference" %in% x),
      mode_class = case_when(
        has_inline & has_ref ~ "Both",
        has_inline ~ "Inline only",
        TRUE ~ "References only"
      )
    )
}

mode_feb <- classify_modes(provider_mode_feb)
mode_dec <- classify_modes(provider_mode_dec)

mode_compare <- bind_rows(
  mode_dec |> mutate(period = "Dec 2025"),
  mode_feb |> mutate(period = "Feb 2026")
) |>
  count(period, mode_class, name = "n_payers") |>
  mutate(
    mode_class = factor(
      mode_class,
      levels = c("References only", "Both", "Inline only")
    ),
    period = factor(period, levels = c("Dec 2025", "Feb 2026"))
  )
```

## Provider references vs. inline groups

The v2 schema eliminates inline provider groups in favor of provider
references, a structural change that reduces file duplication and file size.
In December 2025, **`r sum(mode_dec$mode_class == "Inline only")`** payers
used inline groups exclusively. By February 2026, that dropped to
**`r sum(mode_feb$mode_class == "Inline only")`**.

```{r plot-provider-mode, fig.width=8, fig.height=4.5}
ggplot(
  mode_compare,
  aes(x = mode_class, y = n_payers, fill = period)
) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(
    aes(label = n_payers),
    position = position_dodge(width = 0.7),
    vjust = -0.5,
    size = 4,
    fontface = "bold"
  ) +
  scale_fill_manual(
    values = c("Dec 2025" = "grey65", "Feb 2026" = "steelblue")
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Shift from inline provider groups to provider references",
    subtitle = paste0(
      "Payers using inline-only groups fell from ",
      sum(mode_dec$mode_class == "Inline only"), " to ",
      sum(mode_feb$mode_class == "Inline only"),
      " between December 2025 and February 2026"
    ),
    x = NULL,
    y = "Number of payers",
    fill = NULL,
    caption = paste0(
      "Source: inr_in_network_neg_rates_provider_groups ",
      "and inr_in_network_neg_rates_provider_references\n",
      "Note: Dec 2025 includes ", n_distinct(mode_dec$payer_id),
      " payers; Feb 2026 includes ", n_distinct(mode_feb$payer_id),
      ". Counts differ due to ingestion timing."
    )
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(size = 16, face = "bold", margin = margin(b = 4)),
    plot.subtitle = element_markdown(size = 12, margin = margin(b = 12)),
    plot.caption = element_text(size = 9.5, lineheight = 1.1, margin = margin(t = 12)),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 11),
    legend.position = "top",
    legend.text = element_text(size = 11),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r prep-network-names}
nn_payers <- n_distinct(network_names$payer_id)
total_feb_payers <- n_distinct(payer_feb$payer_id)

# Quality: how many use generic "UNSPECIFIED_NETWORK"?
nn_unspec <- network_names |>
  filter(str_detect(sample_network_name, "(?i)unspecified"))
nn_real <- network_names |>
  filter(!str_detect(sample_network_name, "(?i)unspecified"))
nn_real_payers <- n_distinct(nn_real$payer_id)
```

## Network name adoption

The v2 schema requires a `network_name` field in provider references,
enabling consumers to identify which provider network applies to each rate.
In February 2026, **`r nn_payers`** of `r total_feb_payers` payers
(`r round(nn_payers / total_feb_payers * 100)`%) have at least one provider
reference with a populated `network_name`.

However, quality varies. Two payers populate the field with a generic
`UNSPECIFIED_NETWORK` placeholder. Excluding those,
**`r nn_real_payers`** payers provide meaningful network names, with
**`r comma(n_distinct(nn_real$sample_network_name))`** distinct values
across the dataset.

```{r plot-network-names, fig.width=8, fig.height=5}
# Top network names by frequency of payer/DSN combos
nn_top <- network_names |>
  filter(!str_detect(sample_network_name, "(?i)unspecified")) |>
  count(sample_network_name, name = "n_files", sort = TRUE) |>
  slice_head(n = 15) |>
  mutate(
    sample_network_name = fct_reorder(sample_network_name, n_files)
  )

ggplot(nn_top, aes(x = n_files, y = sample_network_name)) +
  geom_col(fill = "steelblue", width = 0.7) +
  geom_text(
    aes(label = comma(n_files)),
    hjust = -0.15,
    size = 3.5,
    fontface = "bold"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = "Most common network names in February 2026",
    subtitle = paste0(
      nn_real_payers, " payers provide meaningful network names across ",
      comma(n_distinct(nn_real$sample_network_name)), " distinct values"
    ),
    x = "Number of payer/file combinations",
    y = NULL,
    caption = paste0(
      "Source: hive.public_2026_02.inr_provider_references\n",
      "Excludes UNSPECIFIED_NETWORK placeholder values. ",
      "Shows one sampled network name per payer/file combination."
    )
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(size = 16, face = "bold", margin = margin(b = 4)),
    plot.subtitle = element_markdown(size = 12, margin = margin(b = 12)),
    plot.caption = element_text(size = 9.5, lineheight = 1.1, margin = margin(t = 12)),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r prep-ein}
ein_payer <- ein_agg |>
  group_by(payer_id) |>
  summarise(
    has_business_name = any(has_any_business_name),
    .groups = "drop"
  )

ein_total <- nrow(ein_payer)
ein_with <- sum(ein_payer$has_business_name)
```

## EIN business name adoption

The v2 schema requires a `business_name` field alongside EIN records in the
`tin` struct, making it possible to identify providers by name rather than
just tax ID number. Of the **`r ein_total`** payers with EIN data in
February 2026, **`r ein_with`** (`r round(ein_with / ein_total * 100)`%)
include at least one non-empty `business_name`.

```{r plot-ein, fig.width=7, fig.height=4}
ein_plot <- ein_payer |>
  count(has_business_name, name = "n_payers") |>
  mutate(
    label = if_else(has_business_name, "Has business_name", "No business_name"),
    label = factor(label, levels = c("Has business_name", "No business_name"))
  )

ggplot(ein_plot, aes(x = label, y = n_payers)) +
  geom_col(
    aes(fill = has_business_name),
    width = 0.6,
    show.legend = FALSE
  ) +
  geom_text(
    aes(label = paste0(n_payers, " (", round(n_payers / ein_total * 100), "%)")),
    vjust = -0.5,
    size = 4.5,
    fontface = "bold"
  ) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey65")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "EIN business name adoption in February 2026",
    subtitle = paste0(
      ein_with, " of ", ein_total,
      " payers with EIN data include at least one business_name"
    ),
    x = NULL,
    y = "Number of payers",
    caption = paste0(
      "Source: hive.public_2026_02.inr_provider_references_provider_groups\n",
      "Payer counted as adopting if any EIN record has a non-empty business_name."
    )
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(size = 16, face = "bold", margin = margin(b = 4)),
    plot.subtitle = element_markdown(size = 12, margin = margin(b = 12)),
    plot.caption = element_text(size = 9.5, lineheight = 1.1, margin = margin(t = 12)),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 11),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r prep-summary}
# Build a unified payer-level summary across all v2 features
payer_all <- payer_ver_summary |>
  select(payer_id, has_v2) |>
  left_join(
    mode_feb |> select(payer_id, mode_class),
    by = "payer_id"
  ) |>
  left_join(
    network_names |>
      distinct(payer_id) |>
      mutate(has_network_name = TRUE),
    by = "payer_id"
  ) |>
  left_join(
    ein_payer |> rename(has_business_name_val = has_business_name),
    by = "payer_id"
  ) |>
  mutate(
    has_network_name = coalesce(has_network_name, FALSE),
    uses_references = mode_class %in% c("References only", "Both"),
    has_business_name_val = coalesce(has_business_name_val, FALSE)
  )

summary_long <- tibble(
  feature = c(
    "Self-reports v2",
    "Uses provider\nreferences",
    "Has network_name",
    "Has business_name"
  ),
  adopted = c(
    sum(payer_all$has_v2),
    sum(payer_all$uses_references, na.rm = TRUE),
    sum(payer_all$has_network_name),
    sum(payer_all$has_business_name_val)
  ),
  total = n_payers
) |>
  mutate(
    pct = adopted / total * 100,
    feature = factor(feature, levels = rev(feature))
  )
```

## Summary

Across all three v2 structural requirements, adoption is already substantial
in the first month. The chart below shows the share of February 2026 payers
meeting each requirement.

```{r plot-summary, fig.width=8, fig.height=4.5}
ggplot(summary_long, aes(x = pct, y = feature)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(
    aes(label = paste0(adopted, "/", total, " (", round(pct), "%)")),
    hjust = -0.05,
    size = 4,
    fontface = "bold"
  ) +
  scale_x_continuous(
    limits = c(0, 115),
    breaks = seq(0, 100, 25),
    labels = \(x) paste0(x, "%")
  ) +
  labs(
    title = "v2 TiC schema adoption in February 2026",
    subtitle = paste0(
      "Share of ", n_payers, " payers meeting each v2 requirement"
    ),
    x = "Payers adopting",
    y = NULL,
    caption = paste0(
      "Source: hive.public_2026_02 tables\n",
      "Provider references includes payers using references alone or ",
      "alongside inline groups.\n",
      "business_name assessed only for payers with EIN data (",
      ein_total, " of ", n_payers, ")."
    )
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(size = 17, face = "bold", margin = margin(b = 4)),
    plot.subtitle = element_markdown(size = 13, margin = margin(b = 12)),
    plot.caption = element_text(size = 9.5, lineheight = 1.1, margin = margin(t = 14)),
    axis.text.y = element_markdown(size = 12, face = "bold"),
    axis.text.x = element_text(size = 11),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Provider references have the highest adoption rate — most payers were
already using them before the v2 mandate. Network names and EIN business
names, both new v2 fields, show adoption above 50%, though the
`business_name` figure is inflated by the denominator (only payers with EIN
data are counted). The **`r n_v1_only`** payers still on v1-only files will
need to transition before CMS enforcement tightens.
